---
title: "R Course"
author: "Heidi Karlsen"
date: "2024-04-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Topics

-   RMarkdown

-   R operators

-   Correlations

-   Regression models

-   Statistical tests

-   Visualisations in ggplot

```{r, install and load libraries}

#Packages below not previously installed by you, need to be installed before loading them.

#install.packages("<package name>")

library(gapminder)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(MASS)

```

```{r}
head(gapminder, 20)
```

The gapminder dataset provides data about the population, life expectancy and GDP in multiple countries in the world from 1952 to 2007.

## R operator %\>% and statistics

This operator forwards a value, or the result of an expression, into the next function call/expression. See documentation here: <https://cran.r-project.org/web/packages/magrittr/index.html>

It passes the left-hand side data as the first argument to the function on the right-hand side.

### 1. Calculate trends in life expectancy over time

```{r}
gapminder %>%
  group_by(year)%>%
  summarize(average_lifeExp <-mean(lifeExp))
```

### 2. Calculating average life expectancy and gdp per capita in continents in 2007

(gross domestic product divided by the country's total population

```{r}
gapminder %>%
  summarise(average_lifeExp <- mean(lifeExp), average_gdpPercap <- mean(gdpPercap))
```

```{r}
gapminder %>%
  filter(year == 2007)%>%
  summarise(average_lifeExp <- mean(lifeExp), average_gdpPercap <- mean(gdpPercap))
```

### Exercise: 

Calculate average life expectancy and average gdp per capita in continents

-   In the entire dataset

-   For one year only

```{r}
gapminder %>%
  group_by(continent)%>%
  summarise(average_lifeExp = mean(lifeExp), average_gdpPercap = mean(gdpPercap))

```

```{r}
gapminder %>%
  filter(year == 2007)%>%
  group_by(continent)%>%
  summarise(average_lifeExp = mean(lifeExp), average_gdpPercap = mean(gdpPercap))
  
```

### 3. Sorting your data with arrange() 

```{r}
gapminder%>%
  arrange(lifeExp)
```

Life expectancy in Africa in 2007

```{r}
gapminder%>%
  filter(continent =="Africa", year == 2007)%>%
  arrange(lifeExp)
```

```{r}
gapminder%>%
  filter(continent== "Africa", year == 2007)%>%
  arrange(desc(lifeExp))
```

## Regression models 

### 4a. Plotting relation between life expectancy and gdp per capita

```{r}
gapminder%>%
  ggplot(aes(gdpPercap, lifeExp))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)+ #adding a trend line. 
  coord_cartesian(ylim = c(40, 90))
  
  
```

Trend line = fitting a line which follows the data points. We set the method to lm, for linear model, which gives a trend line calculated with a linear regression.

By default, geom_smooth() shows a standard error ribbon, which can be turned off by setting 'se' to FALSE.

### 4.b Improved plot

-   Given the wide range of GDP per capita values, we can use a log scale. It transforms the scale of the x axis to a logarithmic scale with base 10. Thus every point on the x axis represents a tenfold increase from the previous point.

-   Increase point transparency

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm", se = FALSE, color = "black")+
  scale_x_log10()
```

The straight lines of linear regressions are defined by two properties:

-   Intercept = The y value at the point when x is 0. Thus the value of the y axis where the regression line touches 0 on the x axis

-   Slope = The steepness of the line: The amount the y value increases if you increase x by one.

-   The equation of the line is the intercept plus the slope times the x value: y = intercept + slope \* x

-   This trend line is in general pretty close to the data points (although some exceptions) thus, the linear regression seems to be a reasonable fit.

### Correlations

```{r}
cor(gapminder$lifeExp, gapminder$gdpPercap)
```

### Simple linear regression 

Regression models are a class of statistical models that let you explore the relationship between a response variable/dependent variable and some explanatory/independent variables

Regression makes it possible to make thought experiments, hypotheses, predict the dependent variables, given some independent variables

The dependent variable = variable we want to predict

The independent variable = the variables that explain how the dependent variable will change

There are different types of regression analysis: linear and logistic

-   Linear regression: when the dependent variable is numeric, which is the case here

-   Logistic regression: when the variable is logistic (true or false)

-   Both can be simple or multiple - simple when there is only one explanatory/independent variable, as is the case here.

```{r}

regression_gapminder <- lm(lifeExp ~ gdpPercap, data = gapminder)
regression_gapminder
```

```{r}
summary(regression_gapminder)
```

Using these values to predict life expectancy:

```{r}
LifeExp = 53.96 +0.0007649 *10000
LifeExp
```

#### Interpreting the results:

-   Std.Error = Standard error: how different the population mean is likely to be from a sample mean

-   T value = test statistics: how close the relations is between the distribution of our data and the distribution predicted under the null hypothesis

-   Null hypothesis = the hypothesis that there is no correlation between the entities we are testing. it is the pattern in our data \#(i.e., the correlation between variables or difference between groups) divided by the variance in the data (i.e., the standard deviation). See: <https://www.scribbr.com/statistics/test-statistic/>

-   p(\>\|t\|) = p(robability) value: the probability of finding the given t statistic if the null hypothesis were true.

-   We often say that we can assume from a p value of 0.05 or lower that the pattern we have found is statistically significant. A p value of 0.05 means that there are 5 percent chance, if the null hypothesis were true, that we would find a t statistic as strong as the one we have found Here the p value is less than 2.2e-16, thus very small. It is represented in scientific notation. It can also be written as 0.000000000000000022.

-   Residual standard error: an estimate of the standard deviation of the residuals (prediction errors) in a linear regression model: how much the observed values differ from the predicted values. It measures the overall lack of fit of the model to the data. The lower the better.

-   RSE here: 10.49, given the range of life expectancy, this suggests that there is still a substantial amount of variability in life expectancy that isn't captured by gdp per capita.

-   The adjusted R-squared value (0.3403) shows that about 34 % of the variance in life expectancy can be explained by gdp per capita in the model, thus about 65 % is explained by other factors.

-   F-statistic: Measures how well the linear regression model fits the data. It measures the variation between group means. The F-statistic is a ratio of two variances - the variance explained by the regression model (explained variance) and the variance not explained by the model (residual variance).

-   DF = Degrees of freedom, meaning the number of values in a statistic that are free to vary.

### Exercise 

-   Visualise the relation between life expectancy and gdp per capita in 2007

-   Run a simple linear regression on these data

```{r}
gapminder%>%
  filter(year == 2007)%>%
  ggplot(aes(gdpPercap, lifeExp, color = continent))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm", se = FALSE, color = "black")+
  scale_x_log10()
```

```{r}
gapminder_2007 <- gapminder%>%
  filter(year == 2007)
```

```{r}
regression_gapminder_2007 <- lm(lifeExp ~ gdpPercap, data = gapminder_2007)
summary(regression_gapminder_2007)
```

### More plotting

We prepare for running a multiple linear regression. Let's first analyse the relation between life expectancy and gdp per capita, life expectancy and population - and life expectancy and continent by creating scatter plots with trend lines calculated with linear regression.

```{r}

plot1 <- ggplot(gapminder, aes(gdpPercap, lifeExp))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "gdpPercap", y = "lifeExp")+
  scale_x_log10()

plot2 <- ggplot(gapminder, aes(pop,lifeExp))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "pop", y = "lifeExp")+
  scale_x_log10()

plot3 <- ggplot(gapminder, aes(continent, lifeExp))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(x = "continent", y = "lifeExp")

grid.arrange(plot1, plot2, plot3)
```

### Multiple linear regression

```{r}
regression_gapminder2 <- lm(lifeExp~pop + gdpPercap + continent, data = gapminder)
regression_gapminder2
```

```{r}
summary(regression_gapminder2)
```

LifeExp= 47.81+(6.570e−9×pop)+(0.0004495×gdpPercap)+Continent Coefficients

### Interpretation

We see from the R squared value that this model predicts about 58 % of the variance can be explained by the model.

## Statistical tests

Let's use the survey dataset from the MASS package as an examle

```{r}
head(survey)
```

```{r}
summary(survey)
```

### T-tests

Statistical method to analyse whether there are siginificant differences between the means of two groups. We see if the null hypothesis - that there are no significant difference between the mean - can be rejected or not.

```{r}
t.test(Pulse~Sex, data = survey, var.equal = TRUE, paired = FALSE)
```

## Chi-square test

Statistical method to determine whether there is a statistically significant association between categorical variables.

```{r}
Smoke_Excercise = table(survey$Smoke, survey$Exer)
```

```{r}
Smoke_Excercise
```

```{r}
chisq.test(Smoke_Excercise)
```
